UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-15 ms/main.tex
author summary

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference, eigen2006, Hellenthal747}].

>>>>

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\subsection*{Author summary}
The demographic history of human populations features evolving geographic and social barriers to mating. Over time, these barriers have led to varying levels of genetic relatedness among individuals.  This \emph{population structure} is informative about human history, and can have a significant impact on studies of medical genetics. Because population structure depends on myriad demographic forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure. We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction. Using UMAP is computationally efficient and can identify fine-scale population structure in large population datasets. We find it particularly useful to reveal phenotypic variation among genetically related populations, and argue that it is a useful complement  to principal component analysis in primary data visualization.

\subsection*{Introduction}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference, eigen2006, Hellenthal747}].
